---
title: "DS400: Assignment II (R)"
author: "Mohamed Elsherif (Student ID: 6378374)"
date: "2024-11-30"
output: html_document
---

### Directory set up & Libraries Installation

```{r, include=TRUE}
# Load all libraries needed 
library(rstudioapi)
library(readr)
library(data.table)
library(ggplot2)
library(lubridate)
library(tidyr)
library(dplyr)
library(scales)  
library(RColorBrewer)
library(RSQLite)
options(warn = -1) 
```

# Importing Data

### 1.Use relative paths to load the data.

```{r}
# Set working directory to the R md location
library(here)
knitr::opts_chunk$set(root.dir = here::here())
cat("Current working directory is:", getwd(), "\n")
```

### 2. Load the data athletes and medals data into your environment. Print the first 10 observations in each dataset. Also print all variable names for each dataset.

```{r}
# Load athletes and medals datasets 
athletes <- read_csv("./data/athletes.csv")  
medals <- read_csv("./data/medals.csv")  

# Print first 10 rows of each dataset
cat("First 10 observations in the athletes dataset:\n")
head(athletes, 10)

cat("\nFirst 10 observations in the medals dataset:\n")
head(medals, 10)

# Print column names for both datasets
cat("\nColumn names in the athletes dataset:\n")
print(names(athletes))

cat("\nColumn names in the medals dataset:\n")
print(names(medals))

```

### 3. How many different nationalities are there in the athletes dataset?

```{r}
# Remove NA values, then get the unique nationalities
unique_nationalities <- unique(athletes$nationality[!is.na(athletes$nationality)])

# Print number of unique nationalities
cat("Number of different nationalities in the athletes dataset:", length(unique_nationalities), "\n")
```

# Data Visualization

### 4. Create a barplot showing the number of athletes per nationality (from highest to lowest). What is wrong with this plot?

```{r}
# Remove NA values, then get the count of athletes per nationality stored as a table 
athletes_per_nationality <- table(athletes$nationality[!is.na(athletes$nationality)])

# Sort the table from highest to lowest
athletes_per_nationality_sorted <- sort(athletes_per_nationality, decreasing = TRUE)

# plot barplot
barplot(
  athletes_per_nationality_sorted, 
  main = "Number of Athletes per Nationality", 
  ylab = "Number of Athletes", 
  las = 2, 
  col = "lightblue", 
  border = "blue"
)
```

**comment** The barplot has several issues: a large number of categories can clutter the x-axis, especially if NaN or empty nationalities are not filtered out, and the dominance of a few nationalities may overshadow less-represented ones, making them nearly invisible. Additionally, displaying all categories can lead to a top-heavy visualization that lacks meaningful comparison. In R, the barplot() function is not able to fit all the labels on the x-axis by default.

### 5. Make another barplot where only the most common nationalities are displayed (defined as having at least 100 athletes at the Olympics). The remaining athletes should be grouped into a "other" bar. Include the actual number of athletes in each bar.

```{r}
# Identify nationalities with at least 100 athletes
common_nationalities <- athletes_per_nationality[athletes_per_nationality >= 100]

# Get nationalities with less than 100 athletes and group as 'Other'
other_nationalities_count <- sum(athletes_per_nationality[athletes_per_nationality < 100])

# Add 'Other' category to table
common_nationalities <- c(common_nationalities, "Other" = other_nationalities_count)

# Plot 
plotMitNumb <- barplot(
  common_nationalities, 
  main = "Most Common Nationalities with At Least 100 Athletes",
  ylab = "Number of Athletes", 
  las = 2, 
  col = "lightblue", 
  border = "blue",
  names.arg = names(common_nationalities),
  ylim = c(0, max(common_nationalities) * 1.1)
)

# Add number of athletes
text(
  x = plotMitNumb, 
  y = common_nationalities, 
  labels = common_nationalities, 
  pos = 3, 
  cex = 0.8, 
  col = "black"
)
```

### 6. Using the medals dataset, select the top 5 countries with the most medals. Create a plot showing the number of medals on the y-axis and the countries on the x-axis. Color the bars per gender. To make the plot easier to read, include only athletes that identify themselves as male or female. Use a colorblind-friendly palette

```{r}
# Filter medals dataset to include only male and female athletes
medals_filtered <- medals[medals$gender %in% c("M", "W"), ]

# Count total number of medals per country grouped by gender
country_medals_gender <- table(medals_filtered$country_code, medals_filtered$gender)

# Sum total medals for each country
total_medals_by_country <- rowSums(country_medals_gender)

# Sort countries by total medals in descending order then select top 5
top_countries <- names(sort(total_medals_by_country, decreasing = TRUE)[1:5])

# Subset dataset to include only top 5 countries
top_countries_data <- medals_filtered[medals_filtered$country_code %in% top_countries, ]

# Count number of medals per country and gender
top_country_medals_gender <- table(top_countries_data$country_code, top_countries_data$gender)

# convert to a data frame
top_country_medals_gender_df <- as.data.frame(top_country_medals_gender)
colnames(top_country_medals_gender_df) <- c("Country", "Gender", "Medals")

# colorblind palette
colors <- brewer.pal(3, "Set1")  

# plot barplot
ggplot(top_country_medals_gender_df, aes(x = Country, y = Medals, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors[1:2]) +  
  labs(title = "Top 5 Countries by Number of Medals",
       x = "Countries", 
       y = "Number of Medals",
       fill = "Gender") +
  theme_minimal()
```

### 7. Re-create the previous plot but facet by medal type. Are there any differences between the medal types?

```{r}
# Count total number of medals per country grouped by gender + medal type
country_medals_gender_type <- table(medals_filtered$country_code, medals_filtered$gender, medals_filtered$medal_type)

# convert to a data frame
country_medals_gender_type_df <- as.data.frame(country_medals_gender_type)
colnames(country_medals_gender_type_df) <- c("Country", "Gender", "MedalType", "Medals")

# Filter to include only the top 5 countries (based on total medals)
total_medals_by_country <- rowSums(table(medals_filtered$country_code, medals_filtered$gender))
top_countries <- names(sort(total_medals_by_country, decreasing = TRUE)[1:5])

# Subset data to include only top 5 countries
top_countries_data <- country_medals_gender_type_df[country_medals_gender_type_df$Country %in% top_countries, ]

# colorblind palette
colors <- brewer.pal(3, "Set1") 

# plot barplot with faceting by medal type
ggplot(top_countries_data, aes(x = Country, y = Medals, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors[1:2]) +  
  labs(title = "Top 5 Countries by Number of Medals (Faceted by Medal Type)",
       x = "Countries", 
       y = "Number of Medals",
       fill = "Gender") +
  facet_wrap(~MedalType, scales = "free_y") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

**Comment** The USA shows stronger performance for women across all medal types, particularly in Gold and Silver medals. while France and Great Britain show a stronger performance for men in Silver and Bronze medals, though GBR balances Gold medals equally.

### 8. Create a plot that shows the cumulative medals won by the top 3 country over time?

```{r}
# collect total number of medals per country over time
medals_summary <- medals %>%
  filter(!is.na(country_code) & !is.na(medal_date)) %>%  
  group_by(country_code) %>%
  summarise(total_medals = n()) %>%
  arrange(desc(total_medals))  

# Select top 3 countries
top_3_countries <- head(medals_summary$country_code, 3)

# Filter for top 3 countries
medals_top_3 <- medals %>%
  filter(country_code %in% top_3_countries & !is.na(medal_date)) %>%
  mutate(medal_date = as.Date(medal_date))   

# Calculate cumulative medals over time for each country
medals_cumulative <- medals_top_3 %>%
  group_by(country_code, medal_date) %>%
  summarise(daily_medals = n()) %>%  
  arrange(medal_date) %>%
  mutate(cumulative_medals = cumsum(daily_medals))  


ggplot(medals_cumulative, aes(x = medal_date, y = cumulative_medals, color = country_code, group = country_code)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 2) +      
  labs(
    title = "Cumulative Medals Won by the Top 3 Countries Over Time",
    x = "Date",
    y = "Cumulative Medals",
    color = "Country"
  ) +
  scale_x_date(
    date_breaks = "1 day",  
    date_labels = "%b %d"  
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)  
  )
```

### 9. Import the athletics dataset (discipline_athletics.csv). Keep only the data from women's and men's 100m and 200m finals. Print the dimensions of the resulting dataset.

```{r}
# Import Discp. athletics dataset
discipline_athletics <- read_csv("./data/discipline_athletics.csv")

# Filter data to keep only men's and women's 100m and 200m finals
MW_100200 <- discipline_athletics[ 
  discipline_athletics$event_code %in% c('ATHM100M', 'ATHM200M', 'ATHW100M', 'ATHW200M') & 
  discipline_athletics$stage == 'Final' & 
  discipline_athletics$gender %in% c('M', 'W'), 
]

# Print dimensions of Discp. athletics dataset
cat("Dimensions of the filtered dataset:", dim(MW_100200), "\n")
```

### 10. Create two plots (side-by-side) that shows the distribution of the athletes' ages and their finish time, once for the 100m final and once for the 200m final. Color the athletes by their gender. Do not use a shared axes for the two plots. What can you observe from the plots?

```{r}
# Convert date columns to Date format and calculate age
MW_100200$date <- as.Date(MW_100200$date)
athletes$birth_date <- as.Date(athletes$birth_date)

# Merge two datasets on the participant_code and code
merged_data <- merge(MW_100200, athletes, by.x = "participant_code", by.y = "code")

# Calculate age as (difference between the date and birth_date)
merged_data$age <- as.numeric(difftime(merged_data$date, merged_data$birth_date, units = "days")) / 365.25

# Split dataset for 100m and 200m 
data_100m <- merged_data[grepl("ATHM100M|ATHW100M", merged_data$event_code), ]
data_200m <- merged_data[grepl("ATHM200M|ATHW200M", merged_data$event_code), ]

# plot for 100m 
p1 <- ggplot(data_100m, aes(x = age, y = result, color = gender.x)) +
  geom_point() +
  labs(title = "100m Final: Age vs Finish Time", 
       x = "Age (Years)", 
       y = "Finish Time (Seconds)", 
       color = "Gender") +  # Use labs() to set the legend title
  scale_color_manual(values = c("M" = "blue", "W" = "red")) +
  theme_minimal()

# plot for 200m 
p2 <- ggplot(data_200m, aes(x = age, y = result, color = gender.x)) +
  geom_point() +
  labs(title = "200m Final: Age vs Finish Time", 
       x = "Age (Years)", 
       y = "Finish Time (Seconds)", 
       color = "Gender") +  # Use labs() to set the legend title
  scale_color_manual(values = c("M" = "blue", "W" = "red")) +
  theme_minimal()

# put the two plots side by side
library(gridExtra)
grid.arrange(p1, p2, ncol = 2)
```

**comment** In both the 100m and 200m final races, younger athletes, particularly those in their early 20s, tend to perform the best, with male athletes consistently achieving faster times than their female counterparts. The fastest times in both races are dominated by male athletes, with female athletes generally recording slower times and showing more variability in performance. Age also plays a role, as older athletes tend to have slower finish times, especially in the female category. Overall, the 100m and 200m finals shows the advantage of youth and the gender performance disparity in sprinting events

# SQL

### 11. Print the names of all tables in the database.

```{r}
db_path <- "./data/imdb.db"

# Establish connection 
conn <- dbConnect(SQLite(), db_path)

# list all tables in the database
query <- "SELECT name FROM sqlite_master WHERE type='table';"

# Execute query 
tables <- dbGetQuery(conn, query)

cat("Tables in the IMDb database:\n")
print(tables)
```

### 12. Print the first 10 rows of the names table.

```{r}
# select first 10 rows from the 'names' table
query <- "SELECT * FROM names LIMIT 10;"

# Execute query 
names_data <- dbGetQuery(conn, query)

cat("First 10 rows of the 'names' table:\n")
print(names_data)
```

### 13. How many unique names exist in the names table?

```{r}
# count number of unique names in 'names' table
query <- "
SELECT COUNT(DISTINCT primaryName) 
FROM names
WHERE primaryName IS NOT NULL;
"
# Execute query 
unique_names_count <- dbGetQuery(conn, query)

cat("Number of unique names in the 'names' table:", unique_names_count[1,1], "\n")
```
### 14. What is the average rating of all titles in the ratings table?

```{r}
# get column information of 'ratings' table
query <- "PRAGMA table_info(ratings);"

# Execute query 
ratings_columns <- dbGetQuery(conn, query)

cat("Columns in 'ratings' table:\n")
print(ratings_columns)

# calculate average rating from 'ratings' table
query <- "SELECT AVG(averageRating) FROM ratings;"

# Execute query 
avg_rating <- dbGetQuery(conn, query)

cat("Average rating of all titles in 'ratings' table:", avg_rating[1,1], "\n")
```

### 15. What is the average rating of all titles in the ratings table that have less than 100 votes?

```{r}
# calculate average rating for titles with fewer than 100 votes
query <- "
  SELECT AVG(averageRating) AS avg_rating
  FROM ratings
  WHERE numVotes < 100;
"
# Execute query
result <- dbGetQuery(conn, query)

cat("The average rating for titles with less than 100 votes is:", result$avg_rating, "\n")
```

### 16. List the names, birth years, and primary professions of the 10 oldest actors/actresses in the database.

```{r}
# Query to get the 10 oldest actors/actresses
query <- "
  SELECT primaryName, birthYear, primaryProfession
  FROM names
  WHERE (primaryProfession LIKE '%actor%' OR primaryProfession LIKE '%actress%')
    AND birthYear IS NOT NULL
  ORDER BY birthYear ASC
  LIMIT 10;
"
# Execute query
oldest_actors <- dbGetQuery(conn, query)

cat("The 10 oldest actors/actresses:\n")
print(oldest_actors)
```

### 17. Which three movies have the lowest ratings? Print their name, release year, title type, and rating.

```{r}
# find the three movies with lowest ratings
query <- "
  SELECT t.primaryTitle, t.startYear, t.titleType, r.averageRating
  FROM titles t
  JOIN ratings r ON t.tconst = r.tconst
  WHERE t.titleType = 'movie'  
  ORDER BY r.averageRating ASC  
  LIMIT 3;  
"
# Execute query 
lowest_rated_movies <- dbGetQuery(conn, query)

cat("The three movies with the lowest ratings:\n")
print(lowest_rated_movies)
```

### 18. Create a histogram plot that shows the distribution of TV series ratings vs. movie ratings. What can you observe from the plot?

```{r}
# get ratings for movies and TV series
query <- "
  SELECT t.titleType, r.averageRating
  FROM titles t
  JOIN ratings r ON t.tconst = r.tconst
  WHERE t.titleType IN ('movie', 'tvSeries');
"
# Execute query
ratings_data <- dbGetQuery(conn, query)

# Convert title_type to a factor
ratings_data$titleType <- as.factor(ratings_data$titleType)

# plot histogram
ggplot(ratings_data, aes(x = averageRating, fill = titleType)) +
  geom_histogram(position = "identity", alpha = 0.7, bins = 20) +
  scale_fill_manual(values = c("movie" = "blue", "tvSeries" = "red")) +
  labs(
    title = "Distribution of Ratings: Movies vs. TV Series",
    x = "Average Rating",
    y = "Frequency",
    fill = "Title Type"
  ) +
  theme_minimal()
```

**Comment:** **TV Series**are generally rated higher than movies, with more frequent ratings above 7 and fewer very low ratings. On the other hand, **Movies** have a wider distribution of ratings, from extremely low to very high. There also seem to be extreme outliers on both ends for both movies and TV series. Some titles have ratings as low as 1.9 (TV series), and others as high as 9.2 (movies), indicating that both genres have extremes in viewer reception.

### 19. Close the connection to the database (if applicable).

```{r}
dbDisconnect(conn)
```

|                     |
|:--------------------|
| **END OF SOLUTION** |
