---
title: "DS400: Assignment I (R)"
author: "Mohamed Elsherif (Student ID: 6378374)"
date: "2024-11-05"
output: html_document
---

### Directory set up & Libraries Installation

```{r, include=TRUE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
rm(list = ls())
setwd("C:/Users/Admin/Downloads/DS400_WS24-25/Assignment01")
getwd()
install.packages("ggplot2")
options(warn = -1) 
# Load libraries
library(readr)
library(data.table)
library(microbenchmark)
library(ggplot2)
library(lubridate)
library(tidyr)
library(dplyr)
library(stats)
```

# Importing Data

### 1. Reading cvs files using three different functions

```{r}
# Define file paths
listings <- "./data/listings_muc.csv"
calendar <- "./data/calendar_muc.csv"

# Read the listings data using the 3 different methods
AirBNB_listings_muc_utils <- utils::read.csv(listings)
AirBNB_listings_muc_readr <- readr::read_csv(listings, show_col_types = FALSE)
AirBNB_listings_muc_data_table <- data.table::fread(listings)

# Read the calendar data using the 3 different methods
AirBNB_calendar_muc_utils <- utils::read.csv(calendar)
AirBNB_calendar_muc_readr <- readr::read_csv(calendar, show_col_types = FALSE)
AirBNB_calendar_muc_data_table <- data.table::fread(calendar)
```

### 2. Print the first five rows and column types

```{r}
# Print the first five rows and column types for listings data
cat("First five rows of listings data (utils):")
head(AirBNB_listings_muc_utils, 5)
cat("Column types of listings data (utils):")
str(AirBNB_listings_muc_utils)
```

```{r}
# Print the first five rows and column types for calendar data
cat("First five rows of calendar data (utils):")
head(AirBNB_calendar_muc_utils, 5)
cat("Column types of calendar data (utils):")
print(sapply(AirBNB_calendar_muc_utils, class)) 
```

# Profiling the Import

### 3. Computation time of read.csv, read_csv and fread

```{r}
mbm <- microbenchmark(
  listings_utils = read.csv(listings),
  listings_readr = read_csv(listings, show_col_types = FALSE),
  listings_fread = fread(listings),
  times = 10
)
print(mbm)
```

compare the three functions

```{r}
autoplot(mbm)
```

**comment**

-   ***fread** is the fastest and **read.csv** is the slowest. By comparing the mean time for both functions, we see that fread is appr. 5 times faster than the read.csv function.*

# Tidying the data

### 4. Requirements for Tidy data:

-   Each variable forms a column.
-   Each observation forms a row.
-   Each type of observational unit forms a table.

### 5. Evaluating Tidiness of listings and calendar tables:

#### I- listings_muc.csv

**listings_muc.csv is arguably NOT tidy for the following reasons:**

-   **Main Reason** The 'host_id' and 'host_name' columns form a different type of observational unit and are categorical variables that won't change, Therefore we can create a separate table to store this information. This task wasn't explicitly mentioned in the assignment, so I didn't include it as part of my solution below.
-   The enitre 'neighbourhood_group' column and major portion of 'license' column have missing values, which questions the purpose of having both variables included for analysis. To rectify this, we should remove both.
-   Missing values in columns 'last_review' and its current values are stored as an object datatype, requiring conversion to datetime for analysis. However since dates of 'last_review' offers limited value in current analysis unless feature engineered (e.g., into "days since last review"), I opted to drop it from the dataset.
-   Missing values in column 'reviews_per_month' columns when 'number_of_reviews' has value of 0. To rectify this, we can replace those missing values with 0.
-   The column 'id' needs to be converted from float in to integer or object (to avoid issues with decimals). In R, no need to perform such step as R will automatically remove the '.0'.
-   The columns 'name' and 'host_name' (which has 1 missing value) seem irrelevant to analysis and can be removed.

```{r}
# Drop the 'name', 'host_name', 'neighbourhood_group', 'last_review' and 'license' columns
AirBNB_listings_muc_readr<- AirBNB_listings_muc_readr %>%
  select(-name, -host_name, -neighbourhood_group, -last_review, -license)

# Set 'reviews_per_month' to 0 where 'number_of_reviews' is 0
AirBNB_listings_muc_readr <- AirBNB_listings_muc_readr %>%
  mutate(reviews_per_month = ifelse(number_of_reviews == 0, 0, reviews_per_month))
```

#### II- calendar_muc.csv

**calendar_muc.csv is NOT tidy because:**

-   The 'min_max_nights' information are two different variables placed in one column, and each cell has two values. To rectify this, we should separate into two columns: 'minimum_nights' and 'maximum_nights'.
-   The columns 'price' and 'date' need to be converted to float/numeric (after removing the '\$' sign) and datetime format respectively.
-   The column 'listing_id' need to be converted to integer or object (to avoid issues with decimals). In R, no need to perform such step as R will automatically remove the '.0'.
-   The enitre 'adjusted_price' column has missing values, which questions the purpose of having this variable included for analysis. To rectify this, we should remove it.

```{r}
# Split 'min_max_nights' into 'minimum_nights' and 'maximum_nights'
AirBNB_calendar_muc_readr <- AirBNB_calendar_muc_readr %>%
  separate(min_max_nights, into = c("minimum_nights", "maximum_nights"), sep = "-", convert = TRUE)

# Remove the dollar sign and any commas (if present)
AirBNB_calendar_muc_readr$price <- gsub("[^0-9.]", "", AirBNB_calendar_muc_readr$price)

# Convert the cleaned 'price' column from character to numeric
AirBNB_calendar_muc_readr$price <- as.numeric(AirBNB_calendar_muc_readr$price)

# Drop the 'adjusted_price' column from the data frame
AirBNB_calendar_muc_readr <- AirBNB_calendar_muc_readr %>%
  select(-adjusted_price)
```

### 6. Shape of the tables after making the data tidy

```{r}
# Print the first five rows and column types for listings data
cat("First five rows of listings data (utils): \n")
head(AirBNB_listings_muc_readr, 5)
cat("Shape of listings data (utils):")
print(dim(AirBNB_listings_muc_readr))
cat("Column types of listings data (utils):")
str(AirBNB_listings_muc_readr)


# Print the first five rows and column types for calendar data
cat("First five rows of calendar data (utils): \n")
head(AirBNB_calendar_muc_readr, 5)
cat("Shape of calendar data (utils):")
print(dim(AirBNB_calendar_muc_readr))
cat("Column types of calendar data (utils):")
str(AirBNB_calendar_muc_readr)
```

### 7. Change the name of the listing_id column to id

```{r}
# Rename 'listing_id' to 'id' in the AirBNB_calendar_muc_readr_cleaned data frame
AirBNB_calendar_muc_readr <- AirBNB_calendar_muc_readr %>%
  dplyr::rename(id = listing_id)
#names(AirBNB_calendar_muc_readr)[names(AirBNB_calendar_muc_readr) == "listing_id"] <- "id"
```

### 8. Aggregate the calendar data to the listing level

```{r}
# Aggregate the calendar data on 'id'
AirBNB_calendar_aggregated <- AirBNB_calendar_muc_readr %>%
  dplyr::group_by(id) %>%
  dplyr::summarize(
    minimum_nights = max(minimum_nights, na.rm = TRUE),   
    maximum_nights = min(maximum_nights, na.rm = TRUE),    #
    price = max(price),  
    available = sum(available == TRUE),     # Get maximum price (in case there are variations)
    .groups = "drop"                                       # Drop the grouping
  )
```

### 9. Merge the aggregated calendar and the listings data by id. Call the resulting data frame merged and print its shape.

```{r}
# Merge with the listings dataset
merged <- AirBNB_listings_muc_readr %>%
  left_join(AirBNB_calendar_aggregated, by = "id")

cat("Shape of merged data:", dim(merged), "\n")
head(merged, 5)
```

### 10. How many neighborhoods are there in Munich? Print the number of neighborhoods and all their names. How many and which room types are there?

```{r}
# Count the number of unique neighborhoods in the merged dataset
neighborhoods <- unique(merged$neighbourhood)

# Get the number of neighborhoods
num_neighborhoods <- length(neighborhoods)

# Print the number of neighborhoods and their names
cat("Number of neighborhoods in Munich:", num_neighborhoods, "\n")
cat("Neighborhoods:\n")
print(neighborhoods)


# Count the number of unique room types in the merged dataset
room_types <- unique(merged$room_type)

# Get the number of room types
num_room_types <- length(room_types)

# Print the number of room types and their names
cat("Number of room types:", num_room_types, "\n")
cat("Room Types:\n")
print(room_types)

```

### 11. Regress the number of available days on the price and the minimum_nights.

```{r}
# Fit the regression model
model <- lm(available ~ price + minimum_nights, data = merged)
summary(model)
```

**PLEASE SEE PYTHON VERSION FOR MODEL INTERPRETATION**

### 12. Regress the number of available days on the price and the minimum_nights, but include neighborhood and room type as fixed effects.

```{r}
# Convert neighbourhood and room_type to factors if they are not already
merged$neighbourhood <- as.factor(merged$neighbourhood)
merged$room_type <- as.factor(merged$room_type)

# Fit the regression model with fixed effects for neighborhood and room type
model_fixed_effects <- lm(available ~ price + minimum_nights + neighbourhood + room_type, data = merged)
summary(model_fixed_effects)
```

**PLEASE SEE PYTHON VERSION FOR MODEL INTERPRETATION**

|                        |
|:-----------------------|
| **END OF SOLUTION** |
